Strong lensing selection effects

### 2022-07-25 paper/seleff.tex
lens mock code done

@@

\subsubsection{Stellar mass and redshift distribution}

We generated lens galaxies over a finite redshift range, $0.1 < z < 0.9$.
We chose these lower and upper limits because the value of the critical surface mass density $\Sigma_{\mathrm{cs}}$ becomes very large outside of this range, and the fraction of galaxies that act as strong lenses drops substantially as a result.

@@

%After comparing the predicted and observed fundamental plane scatter for different sets of values of $\sigma_{\mathrm{sps}}$, $\sigma_h$ and $\sigma_\gamma$, we settled on three 
%After comparing the predicted and observed fundmantal plane scatter, 
We settled on three different set of intrinsic scatter parameters, as indicated in \Tref{tab:scatter}. We label them the fiducial, the low-scatter and the high-scatter scenarios.
The high-scatter scenario is ruled out both by the fundamental plane and by weak lensing constraints \citep{Son++22}. 
Moreover, our dynamical model is quite simplistic, as it neglects the effects of orbital anisotropy and departures from spherical symmetry, which are additional sources of scatter. 
Nevertheless, we use it in our experiment in order to obtain a more conservative upper limit on the amplitude of the strong lensing bias.
%A model with the same intrinsic scatter parameters and a more realistic 3-dimensional and orbital structure would likely predict a fundamental plane scatter that is at least as large as the observed one.
\begin{table}
\caption{Intrinsic scatter scenarios}
\label{tab:scatter}
\begin{tabular}{lccc}
\hline
\hline
Model name & $\sigma_{\mathrm{sps}}$ & $\sigma_{\mathrm{h}}$ & Predicted FP \\
 & & & scatter \\
\hline
Fiducial & 0.08 & 0.20 & 0.034 \\
Low scatter & 0.05 & 0.10 & 0.021 \\
High scatter & 0.10 & 0.30 & 0.433 \\
\end{tabular}
\end{table}

\subsection{Background sources}\label{ssec:sources}

\subsection{Observations}\label{ssec:obs}

@@

We then generated samples of $z=0.2$ galaxies from the model of section \ref{ssec:lenses} and used the spherical Jeans equation to predict their central velocity dispersion.
The spherical Jeans equation is \citep{B+T87}
\begin{equation}\label{eq:jeans}
\frac{d(\rho_*\sigma_r^2)}{dr} + \frac{\beta(r)}{r}\rho_*\sigma_r^2 = -\rho_*(r)\frac{GM(r)}{r^2},
\end{equation}
where $\rho_*(r)$ is the 3-dimensional distribution of dynamical tracers (i.e. the stars), $\sigma_r$ the radial component of the velocity dispersion, $\beta(r)$ the orbital anisotropy parameter, and $M(r)$ the mass enclosed within a spherical shell of radius $r$. 
We assumed isotropic orbits ($\beta=0$), then integrated \Eref{eq:jeans} to obtain the seeing-convolved, surface brightness-weighted line-of-sight velocity dispersion within the SDSS spectroscopic aperture\footnote{We used Python code available at \url{https://github.com/astrosonnen/spherical_jeans}, for this purpose.}.
Finally, we fitted the fundamental plane relation of \Eref{eq:fpfull} to the observed stellar mass, half-light radius and central velocity dispersion of the mock sample.
We repeated this procedure for different values of the intrinsic scatter parameters $\sigma_{\mathrm{sps}}$, $\sigma_\mathrm{h}$ and $\sigma_\gamma$.

>>>>

\def\zsource{z_{\mathrm{s}}}
\def\zlens{z_{\mathrm{g}}}

@@

\subsubsection{Stellar mass and redshift distribution}\label{ssub:mstarz}

We generated lens galaxies over a finite redshift range, $0.1 < z < 0.7$.
We chose these lower and upper limits because the value of the critical surface mass density $\Sigma_{\mathrm{cs}}$ becomes very large outside of this range, and the fraction of galaxies that act as strong lenses drops substantially as a result.

@@

%After comparing the predicted and observed fundamental plane scatter for different sets of values of $\sigma_{\mathrm{sps}}$, $\sigma_h$ and $\sigma_\gamma$, we settled on three 
%After comparing the predicted and observed fundmantal plane scatter, 
We settled on three different set of intrinsic scatter parameters, as indicated in \Tref{tab:scatter}. We label them the fiducial, the low-scatter and the high-scatter scenarios.
The high-scatter scenario is ruled out both by the fundamental plane and by weak lensing constraints \citep{Son++22}. 
Moreover, our dynamical model is quite simplistic, as it neglects the effects of orbital anisotropy and departures from spherical symmetry, which are additional sources of scatter. 
Nevertheless, we use it in our experiment in order to obtain a more conservative upper limit on the amplitude of the strong lensing bias.
%A model with the same intrinsic scatter parameters and a more realistic 3-dimensional and orbital structure would likely predict a fundamental plane scatter that is at least as large as the observed one.
\begin{table}
\caption{Intrinsic scatter scenarios}
\label{tab:scatter}
\begin{tabular}{lccc}
\hline
\hline
Model name & $\sigma_{\mathrm{sps}}$ & $\sigma_{\mathrm{h}}$ & Predicted FP \\
 & & & scatter \\
\hline
Fiducial & 0.08 & 0.20 & 0.034 \\
Low scatter & 0.05 & 0.10 & 0.021 \\
High scatter & 0.10 & 0.30 & 0.433 \\
\end{tabular}
\notes{
Adopted values of the intrinsic scatter parameters $\sigma_{\mathrm{sps}}$ and $\sigma_{\mathrm{h}}$ in different simulations. For each set of values, the fourth column indicates the scatter around the fundamental plane predicted via the spherical Jeans analysis of Appendix~\ref{sect:appendixb}.
The observed fundamental plane scatter is $0.035$.
}
\end{table}

\subsection{Background sources: galaxies}\label{ssec:sources}

\ale{To be filled by Shun-Sheng}.
We applied a lower limit to the source redshift, by selecting only sources with $z_s > 0.8$.
This ensures that all of the sources lie behind all of the lenses.
A similar cut could be applied to a real survey using photometric redshift, to reduce the incidence of false positives (e.g. arc-like features physically associated with the lens galaxy) in the lens finding phase.
%This selection is meant to simulate a cut on the redshift (possibly photometric) of the arcs, which would help reduce the incidence of false positives during the lens finding phase.
The resulting number density of sources is ZZZ~arcmin$^{-2}$.
We approximated their spatial distribution as uniform in the sky, that is we neglected clustering of the sources.

\subsection{Background sources: quasars}\label{ssec:quasars}

\subsection{Observations}\label{ssub:obs}

For each of the three intrinsic scatter scenarios, we drew a population of foreground galaxies covering XXX square degrees.
The expectation value of the number of galaxies in the corresponding volume, given the redshift and stellar mass cuts described in section \ref{ssub:mstarz}, is YYY.
%Then we generated a population of sources behind the lenses.
%We approximated the foreground galaxies as isolated: for each source, we only modelled the contribution to the lensing field generated by a single lens.
We approximated the foreground galaxies as isolated: when determining whether a galaxy acts as a strong lens, we only modelled the contribution to the lensing signal from the galaxy itself, and neglected that of the environment.
We discuss the possible implications of this approximation in \Sref{sect:discuss}.
For each lens, we determined its caustics relative to the highest source redshift, $\zsource=2.5$. We then placed sources randomly behind the lens. If at least one source fell within a circular region enclosing the caustics, we proceeded to compute its lensed images.

For the extended sources, we produced images with a pixel size of $0.1''$.
We applied a Gaussian PSF with a FWHM of $0.2''$.
Finally, we assumed a background noise level such that an extended source with half-light radius $0.5''$ and an apparent magnitude in the absence of lensing of $\msource=25$ is detected with $S/N=10$.
%This is similar to the background noise expected for the Euclid Wide survey \citep{Sca++22}.
These values are close to those expected for the Euclid Wide survey \citep{Sca++22}.

We applied the lens selection criterion of section \ref{ssec:}

@@

We then generated samples of $z=0.2$ galaxies from the model of section \ref{ssec:lenses} and used the spherical Jeans equation to predict their central velocity dispersion.
The spherical Jeans equation is \citep{B+T87}
\begin{equation}\label{eq:jeans}
\frac{d(\rho_*\sigma_r^2)}{dr} + \frac{\beta(r)}{r}\rho_*\sigma_r^2 = -\rho_*(r)\frac{GM(r)}{r^2},
\end{equation}
where $\rho_*(r)$ is the 3-dimensional distribution of dynamical tracers (i.e. the stars), $\sigma_r$ the radial component of the velocity dispersion, $\beta(r)$ the orbital anisotropy parameter, and $M(r)$ the mass enclosed within a spherical shell of radius $r$. 
We assumed isotropic orbits ($\beta=0$), then integrated \Eref{eq:jeans} to obtain the seeing-convolved, surface brightness-weighted line-of-sight velocity dispersion within the SDSS spectroscopic aperture\footnote{We used Python code available at \url{https://github.com/astrosonnen/spherical_jeans}, for this purpose.}.
Finally, we fitted the fundamental plane relation of \Eref{eq:fpfull} to the observed stellar mass, half-light radius and central velocity dispersion of the mock sample.
We repeated this procedure for different values of the intrinsic scatter parameters $\sigma_{\mathrm{sps}}$, $\sigma_\mathrm{h}$ and $\sigma_\gamma$, then settled on the three scenarios listed in \Tref{tab:scatter}.
